Design and Implementation of Probabilistic Programming Language Anglican

### 2016-11-22 paper/paper.tex
case study completed

\section{Introduction}
For data science practitioners, statistical inference is typically
just one step in a more elaborate analysis workflow. The first stage
of this work involves data acquisition, pre-processing and cleaning.
This is often followed by several iterations of exploratory model
design and testing of inference algorithms. Once a sufficiently
robust statistical model and a corresponding inference algorithm have
been identified, analysis results must be post-processed, visualized,
and in some cases integrated into a wider production system.

@@

Comparisons of Anglican with other implementations of
probabilistic programming languages~\cite{SGG15}\cite[pp.
32--33]{P16} demonstrate that Anglican achieves state-of-the-art
computational efficiency without sacrificing expressiveness.
Anglican language syntax, compilation, invocation, and runtime
support of Anglican queries are discussed in detail in further
sections.
\section{Design Outline}

@@

\section{Language}

@@

\subsection{Core Library}

@@

\subsection{Tail Call}

@@

\section{Macro-based Compilation}

@@

\subsection {The State}

@@

\subsection{Expression Kinds}

@@

\subsubsection{Literals}

@@

\subsubsection{Opaque Expressions}

@@

\subsubsection{General Expressions}

@@

\subsection{Implementation Highlights}

@@

\subsubsection{Continuations}

@@

\subsubsection{Primitive Procedures}

@@

\subsubsection{Probabilistic forms}

@@

\section{Inference Algorithms}

@@

\subsection{The \texttt{infer} function}

@@

\subsection{Internals of an inference algorithm}

@@

\subsection{Addressing of checkpoints}

@@

\section{Definitions and Runtime Library}

@@

\section{Case Study}

@@

\subsection{The deli dilemma}

@@

\subsection{Anglican query}

@@

There are two possibilities, either the same customer visited
the deli twice, or two different customers came to the deli,
one for lunch, one for dinner. We define an Anglican function
for each case. Note that this is the first time we switch from
Clojure to Anglican.

@@

We are finally ready to define the query.

@@

Performing inference on query \texttt{deli} should predict the
probability that the same customer visited the deli twice, as
well as probability distributions of average arrival times for
both cases.

\subsection{Inference}

@@

\begin{lstlisting}[style=default]
(def N 1000)
(def predicts (map get-predicts
                   (take N (drop N samples))))
\end{lstlisting}

@@

@@

@@

@@

\begin{figure}
    \includegraphics[trim={12pt 0 0 0},scale=0.7]{same-customer-time.pdf}
\end{figure}

\begin{figure}
    \includegraphics[trim={12pt 0 0 0},scale=0.7]{different-customers-times.pdf}
\end{figure}

@@

\section{Summary}

>>>>

\section{Introduction}
\label{sec:intro}

For data science practitioners, statistical inference is typically
just one step in a more elaborate analysis workflow. The first stage
of this work involves data acquisition, pre-processing and cleaning.
This is often followed by several iterations of exploratory model
design and testing of inference algorithms. Once a sufficiently
robust statistical model and a corresponding inference algorithm have
been identified, analysis results must be post-processed, visualized,
and in some cases integrated into a wider production system.

@@

Comparisons of Anglican with other implementations of
probabilistic programming languages~\cite{SGG15}\cite[pp.
32--33]{P16} demonstrate that Anglican achieves state-of-the-art
computational efficiency without sacrificing expressiveness.
Anglican language syntax, compilation, invocation, and runtime
support of Anglican queries are discussed in detail in further
sections.

\section{Design Outline}
\label{sec:design}

@@

\section{Language}
\label{sec:language}

@@

\subsection{Core Library}
\label{sec:core}

@@

\subsection{Tail Call}
\label{sec:tail}

@@

\section{Macro-based Compilation}
\label{sec:compilation}

@@

\subsection {The State}
\label{sec:state}

@@

\subsection{Expression Kinds}
\label{sec:expr}

@@

\subsubsection{Literals}
\label{sec:literals}

@@

\subsubsection{Opaque Expressions}
\label{sec:opaque}

@@

\subsubsection{General Expressions}
\label{seq:general}

@@

\subsection{Implementation Highlights}
\label{seq:highlights}

@@

\subsubsection{Continuations}
\label{seq:continuations}

@@

\subsubsection{Primitive Procedures}
\label{seq:primitive}

@@

\subsubsection{Probabilistic Forms}
\label{seq:forms}

@@

\section{Inference Algorithms}
\label{sec:inference}

@@

\subsection{The \texttt{infer} Function}
\label{sec:infer}

@@

\subsection{Internals of an Inference Algorithm}
\label{seq:internals}

@@

\subsection{Addressing of Checkpoints}
\label{sec:addressing}

@@

\section{Definitions and Runtime Library}
\label{seq:runtime}

@@

\section{Case Study}
\label{seq:study}

@@

The program presented in this section is intentionally short and
simple. Aglican is capable of compiling and running elaborated
programs handling large amounts of data. Advanced examples of
Anglican programs and inference can be found in literature on
applications of probabilistic programming~\cite{PLW15, P16, MPT+16}.

\subsection{The Deli Dilemma}
\label{seq:deli}

@@

\subsection{Anglican Query}
\label{seq:deli-query}

@@

There are two possibilities: either the same customer visited
the deli twice, or two different customers came to the deli,
one for lunch, one for dinner. We define an \textbf{Anglican} function
for each case. Note that this is the first time we switch from
Clojure to Anglican. The functions must be written in Anglican
(and hence defined using \texttt{defm} instead of \texttt{defn})
because they contain probabilistic forms \texttt{sample} and
\texttt{observe}.

@@

We are finally ready to define the query in \textbf{Anglican}.

@@

Performing inference on query \texttt{deli} predicts the
probability that the same customer visited the deli twice, as
well as probability distributions of average arrival times for
both cases.

\subsection{Inference}
\label{seq:deli-inference}

Having defined the query, we are now ready to run the query
using an inference algorithm. Function \texttt{doquery} provided
by the \texttt{anglican.core} namespace accepts the inference
algorithm, the query, and optional parameters, and returns a
lazy sequence of samples. We use here the inference algorithm
called Lightweight Metropolis-Hastings (LMH). The algorithm is
somewhat slow to converge but can be used with any Anglican
query, and should be robust enough for our simple problem.

We bind the results of \texttt{doquery} to variable
\texttt{samples}, to analyse the results later. However, since
the sequence is lazy, no inference is performed and no samples
are generated until they are retrieved and processed.

@@

To approximate the inferred distribution, we extract a finite
subsequence of samples. Many algorithms use an initial
subsequence of samples to converge to the target distribution,
hence we drop initial $N$ samples ($N$ is 5000 in the code
snippet below), and collect the next $N$ samples.

\begin{lstlisting}[style=default]
(def N 5000)
(def predicts (map get-predicts
                   (take N (drop N samples))))
\end{lstlisting}

Based on the collected samples we compute an approximation of
the posterior probability \texttt{p-same+} that the same
customer visited the deli twice.

@@

With the specified observations, \texttt{-same+} is $\approx
0.12$. The results may vary from run to run, and, for given
algorithm, the accuracy depends on the number of samples we
chose to retrieve. The probability is much lower than the
waiter's guess \texttt{p-same} ($\frac 2 3$).

In addition to computing the posterior probability that the same
customer visited the deli twice, we may want to know the average
time (or times) to arrive. In Bayesian inference, it is common
to report distributions instead of `most likely' values. We use
predicted values in retrieved samples to approximate the
distributions, and plot distribution histograms for the same
customer visiting twice (Figure~\ref{fig:dist-same}) and
two different customers (Figure~\ref{fig:dist-diff})

@@

In addition to plotting the distribution histograms, we use
functions \texttt{mean}, \texttt{std} provided, along with other
useful statistical functions, in the \texttt{anglican.{\linebreak[0]}stat}
namespace. This is another illustration of advantage of tight
integration between Clojure and Anglican --- probabilistic
models are expressed in Anglican, however processing of data and
results can rely on the full power of Clojure.

\begin{figure}
    \includegraphics[trim={12pt 0 0 0},scale=0.7]{same-customer-time.pdf}
    \caption{Arrival time distribution for a single customer.}
    \label{fig:dist-same}
\end{figure}

\begin{figure}
    \includegraphics[trim={12pt 0 0 0},scale=0.7]{different-customers-times.pdf}
    \caption{Arrival time distributions for two customers.}
    \label{fig:dist-diff}
\end{figure}

For inference, we chose to use  Lightweight
Metropolis-Hastings, perhaps somewhat voluntarily. A strength of
probabilistic programming is that models are separated from
inference. To switch to a different inference algorithm we just
need to pass a different value to \texttt{doquery}. For example,
we may decide to use Black-Box Variational Bayes (BBVB), which
may not work equally well for all probabilistic programs, but is
much faster to converge.

@@

We can still use samples and summary statistics with BBVB to
approximate the posterior distribution, however variational
inferences approximates the posterior by known distributions,
and we can directly retrieve the distribution parameters.

@@

@@

We can guess that the variational distributions correspond to the
prior distributions in the \texttt{sample} forms, in the order
of appearance. However, it would help if we could use more
informative labels instead of automatically generated symbols
\texttt{S28209}, \texttt{S28217}, etc. Here the option to
specify identifiers for probabilistic forms explicitly comes
handy. If we modify the \texttt{sample} forms to use explicit identifiers
(only the forms are shown for brevity), the output becomes much
easier to analyse.

\begin{lstlisting}[style=default]
          (sample :arrival-time-same
                  time-to-arrive-prior)]

          (sample :arrival-time-first
                  time-to-arrive-prior)

          (sample :arrival-time-second
                  time-to-arrive-prior)]

  (let [is-same-customer (sample :same-or-different
                                 (flip p-same))
\end{lstlisting}

The output becomes much easier to interpret and analyse
programmatically.

\begin{verbatim}
{:arrivate-time-same
 {(0 anglican.runtime.normal-distribution)
  {:mean 10.99753360180663,
   :sd 0.7290976433082352}},
 :arrival-time-first
 {(0 anglican.runtime.normal-distribution)
  {:mean 12.668050292254202,
   :sd 0.9446695174790353}},
 :arrival-time-second
 {(0 anglican.runtime.normal-distribution)
  {:mean 9.104132559955836,
   :sd 0.9479290526821788}},
 :same-or-different
 {(0 anglican.runtime.flip-distribution)
  {:p 0.11671977016875924,
   :dist
   {:min 0.0,
    :max 1.0}}}}
\end{verbatim}

This completes the case study, where we showed a probabilistic
programming solution of a problem, implemented and analysed in
Anglican and Clojure, using two different inference algorithms.

\section{Summary}
\label{seq:summary}
